Supplement for Algorithms for Estimating Trends in Global Temperature Volatility

### 2018-09-06 aaai-2018/aaai-paper-supplement.tex
aaai submitted version

%File: formatting-instruction.tex
\documentclass[letterpaper]{article} %DO NOT CHANGE THIS
\usepackage{aaai19}  %Required
\usepackage{times}  %Required
\usepackage{helvet}  %Required
\usepackage{courier}  %Required
\usepackage{url}  %Required
\usepackage{graphicx}  %Required
\frenchspacing  %Required
\setlength{\pdfpagewidth}{8.5in}  %Required
\setlength{\pdfpageheight}{11in}  %Required
%PDF Info Is Required:

\pdfinfo{
  /Title (Algorithms for Estimating Trends in Global Temperature Volatility)
   /Author (Khodadadi and McDonald)
  /Author (John Doe, Jane Doe)}

@@

\usepackage{gensymb}
\usepackage{xcolor}
\newcommand{\attn}[1]{\textcolor{red}{TODO: #1}}
\newcommand{\one}{\mathbf{1}}
\renewcommand{\algorithmiccomment}[1]{\hfill $\rhd$ #1}
\newcommand{\given}{\;\vert\;}
\newcommand*{\algorithmautorefname}{Algorithm}%
\newcommand{\norm}[1]{\left\lVert #1 \right\rVert}
\DeclareMathOperator*{\prox}{\bf prox}
\DeclareMathOperator*{\sign}{sign}

\usepackage{xr}
\externaldocument{aaai-paper-supplement}

@@

\title{Algorithms for Estimating Trends in Global Temperature Volatility}

@@

\begin{abstract}
Trends in terrestrial temperature variability are perhaps more relevant for species
viability than trends in mean
temperature. In this paper, we develop methodology for estimating such
trends using multi-resolution climate data from polar orbiting weather
satellites. We derive two novel algorithms for computation that are
tailored for dense, gridded observations over both 
space and time. We evaluate our methods with a simulation constructed
to mimic these data's 
features and on a large, publicly available, global
temperature dataset with the eventual goal of tracking trends in
cloud reflectance temperature variability.
\end{abstract}

\section{Introduction}

% Estimating smooth trends over time for large collections of time
% series is a common problem in economics, finance, meteorology,
% neuroscience and more. Most of this work focuses on analyzing trends
% (or removing trends) in the temporal average, but trends in variance
% can be more relevant, especially for financial data but also for
% climate science.

The amount of sunlight reflected from clouds is among the largest
sources of uncertainty in climate
prediction~\citep{BoucherRandall2013}. But climate models fail to
reproduce global cloud statistics, and understanding the reasons for
this failure is a grand challenge of the World Climate Research
Programme~\citep{BonyStevens2015}.
% By understanding how cloudliness changes over time, examining changes
% in brightness temperature within various cloud types, and inferring
% changes in cloud radiative effects, atmospheric scientists can better model climate
% change. Modeling inter-satellite biases, satellite drift, and
% seasonal and long-term climate phenomena like El Ni\~{n}o will lead to
% a better understanding of climate
% change~\citep{SchreierKahn2014,BaumWielicki1994,BaumWielicki1992,FreyAckerman1996}.
While numerous
studies have examined the overall impacts of clouds on climate
variability~\citep{MyersMechoso2018,GrisePolvani2013,BenderRamanathan2012},
such investigations have been hampered by the lack of a suitable
dataset. Ideal data would have global coverage at high spatial
resolution, a long enough record to recover temporal trends, and be
multispectral~\citep{WielickiYoung2013}. 
% The International Satellite
% Cloud Climatology Project (ISCCP)
% has been producing cloud property information from
% limited spectral channels for over three
% decades~\citep{RossowSchiffer1991}. These data have been used in 
% various climate variability
% studies~\citep[e.g.]{BenderRamanathan2012}, however, the utility of
% these data for long-term study of climate variability has been
% questioned~\citep{EvanHeidinger2007}. 
% ISCCP provides a relatively long record, but it only incorporates
% radiances from limited visible and infrared channels from Advanced
% Very High Resolution Radiometer (AVHRR) instruments. 
Current work~\cite{StatenKahn2016,SchreierKahn2010,KahnFishbein2007} seeks to create a
spectrally-detailed dataset by %which can avoid the above issues by
combining radiance data from Advanced Very High Resolution Radiometer
imagers with readings from High-resolution 
Infrared Radiation Sounders on board legacy
satellites. 
In anticipation of this new dataset, our work develops novel methodology
for examining the trends in variability of climate data across space
and time.

\subsection{Variability Rather Than Average}
\label{sec:vari-rath-than}

Trends in terrestrial temperature variability are perhaps more relevant for species
viability than trends in mean
temperature~\cite{huntingford_no_2013},
because an increase in
temperature variability will increase the probability of extreme hot
outliers~\cite{VasseurDeLong2014}. Recent climate literature
suggests that it is more difficult for society to adapt to these
extremes than to the gradual increase in the mean temperature
\cite{hansen_perception_2012,huntingford_no_2013}. Furthermore, the willingness of popular media to
emphasize the prevalence extreme cold events coupled with a
fundamental misunderstanding of the relationship between climate (the global
distribution of weather over the long run) and weather (observed
short-term, localized behavior) leads to public misunderstanding of climate
change. In fact, a point of active debate is the extent to which the observed
increased frequency of extreme cold events in the 
northern hemisphere can be attributed to increases in temperature
variance  rather than to increases in
mean climate~\cite{Screen2014,FischerBeyerle2013,TrenberthZhang2014}.

Nevertheless, research examining trends in the volatility of
spatio-temporal climate data is scarce. \citet{hansen_perception_2012} studied the change in
the standard deviation (SD) of the surface temperature in the NASA
Goddard Institute for Space Studies gridded temperature data set by
examining the empirical SD at each spatial location relative to that
location's SD over a base period, and showed
that these estimates are increasing.
\citet{huntingford_no_2013} took a similar
approach in analyzing the ERA-40 data set. They argued that, while there
is an increase in the SDs from 1958-1970 to 1991-2001, it
is much smaller than found in
\cite{hansen_perception_2012}. \citet{huntingford_no_2013} also computed the
time-evolving global SD from the de-trended time-series at each
position and argued that the global SD has been
stable.

These and other related work  
\cite[e.g.,]{rhines_frequent_2013} have several shortcomings. First, no
statistical analysis has been performed to examine if the changes in
the SD are statistically significant. Second, the methodologies for
computing the SDs are highly sensitive to the choice of base period. Third,
and most importantly, temporal and spatial correlations between 
observations are completely ignored.

In the present work, we examine variance (rather than the mean) for a number of reasons.
First, instrument bias in the satellites increases over time so
examining the mean over time conflates that bias with any actual
change in mean (though the variance is unaffected). 
Second, extreme weather events (hurricanes, droughts, wildfires in
California, heatwaves in Europe) may be driven more strongly by increases
in variance than by increases in mean. 
Finally, even if the global mean temperature is constant, there may still
be climate change. In fact, atmospheric physics suggests that,
across space, average temperatures should not change (extreme cold
in one location is offset by heat in another). But if swings across
space are becoming more rapid, then, even with no change in mean global
temperature over time, increasing variance can lead to
increases in the prevalence of extreme events.

\subsection{Main Contributions}

The main contribution of this work is to develop a new methodology for
detecting the trend in the volatility of spatio-temporal data. In this
methodology, the variance at each position and time are estimated by
maximizing penalized likelihood
maximization. Following~\cite{Tibshirani2014}, we penalize the 
differences between the estimated variances which
are temporally and spatially ``close'', resulting
in a generalized LASSO problem. However, in our application, the
dimension of this optimization problem is massive, so the
standard solvers are inadequate.

We develop two algorithms
which are computationally feasible on extremely large data. In the
first method, we adopt an optimization technique 
called alternating direction method of multipliers
(ADMM, \citealp{boyd_distributed_2011}), to divide the total problem into
several sub-problems of much lower dimension and show how the total
problem can be solved by iteratively solving these sub-problems. The
second method, called linearized ADMM
\cite{parikh_proximal_2014}, solves the main problem by iteratively
solving a linearized version of it. We will compare the benefits of
each method.

Our main contributions are as follows:
\begin{enumerate}
\item We propose a model for nonparametric variance estimation for a
  spatio-temporal process and discuss the relationship between our
  methods and those existing in the machine learning literature (\autoref{sec:ell_1-trend-filt}).
\item We derive two alternating direction method of multiplier
  algorithms (ADMM) to fit our estimator when applied to 
  very large data (\autoref{sec:prop-optim-meth}). We give situations
  under which each algorithm is most likely to be useful. Open-source
  Python code is available on {\tt github}.
\item Because the construction of satellite-based datasets is ongoing
  and currently proprietary, we illustrate our methods on a large,
  publicly available, global temperature dataset.
  The goal is to demonstrate the feasibility of these methods for tracking
  world-wide trends in variance in standard atmospheric data and a
  simulation constructed to mimic these data's features
  (\autoref{sec:empirical-evaluation}). 
\end{enumerate}

While the motivation for our methodology is its application to large, gridded climate
data, we note that our algorithms are also applicable to neuroimaging,
image denoising, or examining large collections of financial instruments.

\section{Smooth Spatio-temporal Variance Estimation}
\label{sec:ell_1-trend-filt}

\citet{KimKoh2009} proposed $\ell_1$-trend filtering as a
method for estimating a smooth, time-varying trend. It is formulated
as the optimization problem 
% : a time-series $y_t$, $t=1,...,T$ is observed and we believe that
% it consists of a smooth trend and a stochastic component and the
% goal is to estimate the trend. The estimated trend is desired to be
% smooth, but at the same time, we seek small estimated residuals
% (stochastic component). These two objectives cannot be achieved
% simultaneously and so a trade-off between them should be made. In
% $\ell_1$-trend filtering, this trade-off is formulated as the
% following convex optimization problem: 
$$
\min_{\beta} \frac{1}{2} \sum_{t=1}^{T} (y_t-\beta_t)^2+\lambda
\sum_{t=1}^{T-2} \left|\beta_t-2\beta_{t+1}+\beta_{t+2}\right| 
$$
or equivalently:
\begin{equation}
\min_{\beta} \frac{1}{2} \norm{ y-\beta }_2^2+\lambda \norm{ D_t \beta}_1
\label{eq:l1tf}
\end{equation}
 where $y=\{y_t\}_{t=1}^T$ is an observed time-series, $\beta\in \mathbb{R}^T$ is the smooth trend,
 $D_t$ is a $(T-2)\times T$ matrix (see Appendix C), and $\lambda$ is a tuning parameter
 which balances fidelity to the data (small errors in the first term)
 with a desire for smoothness.  
% The \textit{penalty matrix} is:
% \begin{equation}
% D=\begin{bmatrix}
% 1 & -2 & 1 &  &  &  &\\ 
% & 1 & -2 & 1 &  &  &\\ 
% &  & \ddots & \ddots & \ddots  &  &\\ 
% &  & & 1 & -2 & 1 &  \\ 
% &  &  &   & 1 & -2 & 1 
% \end{bmatrix}
% \label{eq:d_matrix}
% \end{equation}
%The first term in \eqref{eq:l1tf} penalizes large residuals while the second term encourages smooth estimated $\beta$.
% In Appendix A, we specify the form of the matrix $D$ which results in piecewise
% linear estimated $\beta$. 
\citet{KimKoh2009} proposed a specialized primal-dual
interior point (PDIP) algorithm for solving \eqref{eq:l1tf}. From a
statistical perspective, \eqref{eq:l1tf} can be viewed as a constrained maximum
likelihood problem with independent observations from a normal
distribution with common variance, $y_t \sim \mbox{N}(\beta_t,
\sigma^2)$, subject to a piecewise linear constraint on
$\beta$. Alternatively, solutions to \eqref{eq:l1tf} are %equivalent to
maximum a posteriori Bayesian estimators based on Gaussian likelihood
with a special Laplace prior distribution on $\beta$. Note that the
structure of the estimator is determined by the penalty function
$\lambda\norm{D_t\beta}_1$ rather than any parametric trend
assumptions---autoregressive, moving average, sinusoidal seasonal
component, etc. The resulting trend is therefore essentially
nonparametric in the same way that splines are nonparametric. In fact,
using squared $\ell_2$ norm as the penalty instead of $\ell_1$ results exactly in
regression splines.

% The results of applying $\ell_1$-trend filtering on the Bloomington
% time-series are shown in \autoref{fig:bloom_detrended}. The left
% panel shows the original time-series (blue) together with the
% estimated $\beta$ (red) and the right panel shows the residuals
% $y_t-\beta_t$ (blue). As it can be seen, the estimated $\beta$s are
% piecewise linear. The number of linear segments is determined by the
% value of $\lambda$. Here, we chose $\lambda=500$. The choice of
% $\lambda$ is a model selection problem and we will return to it in
% the next section ???.

\subsection{Modifications for Variance}
\label{sec:l1tf_var}

Inspired by the $\ell_1$-trend filtering algorithm, we propose a
non-parametric model for estimating the variance of a time-series. To
this end, we assume that at each time step $t$, there is a parameter
$h_t$ such that the observations $y_t$
are independent normal variables with zero mean and variance
$\exp(h_t)$. The negative log-likelihood of the observed data in this
model is $l(y\given h) \propto -\sum_{t=1}^T h_t - y_t^2e^{-h_t}$. Crucially,
we assume that the parameters $h_t$ vary smoothly and estimate them by
minimizing the penalized, negative
log-likelihood:  
\begin{equation}
\min_h -l(y\given h)+\lambda \norm{ D_th }_1
\label{eq:l1tf_var}
\end{equation}
where $D_t$ has the same structure as above.

As with~\eqref{eq:l1tf}, one can solve~\eqref{eq:l1tf_var} using the
PDIP algorithm (as in, e.g.,
\texttt{cvxopt}, \citealp{andersen_cvxopt:_2013}). In each iteration of
PDIP we need to compute a search direction by taking a Newton
step on a system of nonlinear equations. For completeness, we
provide the details in Appendix A of the Supplement, where we show how to
derive the dual of this optimization problem and compute the first and
second derivatives of the dual objective function.

\begin{comment}
\begin{algorithm}[tb]
  \caption{PDIP for $\ell_1$ variance estimation}
  \label{alg:pdip}
  \begin{algorithmic}
    \REQUIRE $\lambda>0$, $w>0$, $\nu\leftarrow 0$, $\mu_1\leftarrow
    0$, $\mu_2\leftarrow 0$, $J\in\mathbb{Z}^+$, $\{w_k\}$
    \COMMENT{Initialization} 
    \FOR[Central path]{$k=1,2,\ldots$}
    \FOR[Newton updates]{$j=1,\ldots,J$}
    \STATE Solve $A [\Delta\nu\; \Delta\mu_1\;
    \Delta\mu_2]^\top=r_{w_k}$ to find the search direction. 
    \STATE $\quad\quad A$ is given in \ref{eq:delta_r}
    \STATE Update $[\nu^{j+1}\; \mu^{j+1}_1\; \mu^{j+1}_2] \leftarrow
    [\nu^{j}\; \mu^{j}_1\; \mu^{j}_2] + [\Delta\nu\; \Delta\mu_1\;
    \Delta\mu_2]$. 
    \ENDFOR
   \ENDFOR
   \RETURN $h=\log\frac{y^2}{1+D^\top\nu}$ 
   \\
   \attn{Is this right? Explicit form of $A$}\\
   \attn{Arash: see appendix A}
  \end{algorithmic}
\end{algorithm}
\end{comment}

\subsection{Adding Spatial Constraints}
\label{sec:exten}

The method in the previous section can be used to estimate the
variance of a single time-series. Here we extend this
method to the case of spatio-temporal data.

At a specific time $t$, the data are measured on a grid of points with
$n_r$ rows and $n_c$ columns for a total of $S=n_r\times n_c$ spatial
locations. Let $y_{ijt}$ denote the value of the 
observation at time $t$ on the $i^\text{th}$ row and $j^\text{th}$
column of the grid, and $h_{ijt}$ denote the corresponding
parameter. We seek to impose both temporal and spatial smoothness 
constraints on the hidden variables. Specifically, we seek a solution
for $h$ which is piecewise linear in time and piecewise constant in
space (although higher-order smoothness can be imposed with minimal
alterations to the methodology). We achieve this goal
by solving the following optimization problem: 
{\scriptsize
\begin{equation}
\label{eq:l1tf_var_st}.
\begin{aligned}
&\min_h \sum_{i,j,t}h_{ijt}+y_{ijt}^2e^{-h_{ijt}} +\lambda_t
\sum_{i,j} \sum_{t=1}^{T-2}
\left|h_{ijt}-2h_{ij(t+1)}+h_{ij(t+2)}\right|\\ 
&+\lambda_s \sum_{t,j} \sum_{i=1}^{n_r-1} \left|h_{ijt}-h_{(i+1)jt}\right|
+\lambda_s \sum_{t,i} \sum_{j=1}^{n_c-1} \left|h_{ijt}-h_{i(j+1)t}\right|
\end{aligned}
\end{equation}
}%

The first term in the objective is proportional to the negative
log-likelihood, the second is the temporal penalty for the
time-series at each location $(i,j)$, while the third and fourth,
penalize the difference between the estimated variance of two
vertically and horizontally adjacent points, respectively. The spatial
component of this
penalty is a special case of trend filtering on
graphs~\cite{WangSharpnack2016} which penalizes the difference between
the estimated values of the signal on the connected nodes (though the
likelihood is different). As before,
we can write \eqref{eq:l1tf_var_st} in matrix form where $h$ is a
vector of length
$TS$ and $D_t$ is replaced by $D \in
\mathbb{R}^{(N_t+N_s) \times (T \cdot S)	}$ (see Appendix C), where $N_t=S \cdot
(T-2)$ and $N_s=T \cdot (2n_rn_c-n_r)$ are the number of temporal and
spatial constraints, respectively. 
% The exact form of this matrix in clarified in Appendix
% B in the Supplement. 
Then, as we have two different tuning parameters
for the temporal and spatial components, we write $\Lambda
=\left[\lambda_t\one_{N_t}^\top,\;
  \lambda_s\one_{N_s}^\top\right]^\top$ leading
to:\footnote{Throughout the paper, we use $|x|$ for both scalars and
  vectors. For vectors we use this to denote a vector obtained by
  taking the absolute value of each entry of $x$.}  
% let $h$ be a
% vector whose first $T$ entries are $h_{11t}$ for $t=1,...,T$, its next
% $T$ entries are $h_{21t}$ and so on, then the optimization problem in
% the matrix form is as
\begin{equation}
\min_h -l(y\given h)+ \Lambda^\top | Dh |.
\label{eq:l1tf_var_st_mat}
\end{equation}

\subsection{Related Work}

Variance estimation for financial time series has a lengthy history,
focused especially on parametric models like the generalized
autoregressive conditional heteroskedasticity (GARCH) process~\cite{engle2002dynamic} and
stochastic volatility models~\cite{HarveyRuiz1994}. These models (and
related AR processes) are specifically for parametric modelling of
short ``bursts'' of high volatility, behavior typical of financial
instruments. Parametric models for spatial data go back at least
to~\cite{besag1974spatial} who proposed a conditional probability
model on the lattice for examining plant ecology.

More recently, nonparametric models for both spatial and temporal data
have focused on using $\ell_1$-regularization for trend
estimation. \citet{KimKoh2009} proposed $\ell_1$-trend filtering for
univariate time series, which forms the basis of our methods. These
methods have been generalized to higher order temporal smoothness
~\cite{Tibshirani2014}, graph dependencies~\cite{WangSharpnack2016},
and, most recently, small, time-varying
graphs~\cite{HallacPark2017}.

Our methodology is similar in flavor to~\cite{HallacPark2017} or
related work in~\citep{GibberdNelson2017,MontiHellyer2014}, but with
several fundamental differences. These papers aim to
discover the time-varying structure of a network. To achieve this
goal, they use Gaussian likelihood with unknown precision matrix and
introduce penalty terms which (1) encourage sparsity among the
off-diagonal elements and (2) discourage changes in the estimated inverse
covariance matrix from one time-step to the next.
Our goal in the present work is to detect the temporal trend in the variance of each
point in the network, but the network is known (corresponding to the grid
over the earth). To modify the
objective function in~\cite[Eq.\ 2]{HallacPark2017}, we would
enforce complete sparsity on the off-diagonal elements (since
they are not estimated) and add a new penalty to enforce spatial
behavior across the diagonal elements. Thus, 
\eqref{eq:l1tf_var_st_mat} is not simply a special case of these
existing methods. Finally, these papers examine networks with hundreds
of nodes and dozens to hundreds of time points. As discussed next, our data are
significantly larger than these networks and attempting to estimate a
full covariance would be prohibitive, were it required.

\section{Optimization Methods}
\label{sec:prop-optim-meth}

For a spatial grid of size $S$ and $T$ time steps, $D$ in equation \ref{eq:l1tf_var_st_mat} will have
$3Tn_rn_c-2n_rn_c-Tn_r$ rows and $S \cdot T$ columns. For a $1^\circ\times
1^\circ$ grid over the entire northern hemisphere and daily data over
10 years, we have $S=90\times 360\approx 32,000$ spatial locations and
$T=3650$ time points, so $D$ has
approximately $10^8$ columns and $10^8$ rows. In principal, we could
solve~\eqref{eq:l1tf_var_st_mat} using PDIP as before, however, each iteration
requires solving a linear system of equations which
depends on $D^\top D$. Therefore,
applying the PDIP directly is infeasible for our data.\footnote{We
  note that $D$ is a highly structured, sparse matrix, but, unlike
  trend filtering alone, it is not banded. We are unaware of general
  linear algebra techniques for inverting such matrix, despite our
  best efforts to find them.}

In the next section, we develop two ADMM algorithms for solving this
problem efficiently. The first casts the problem as a
so-called consensus optimization problem~\cite{boyd_distributed_2011}
which solves smaller sub-problems using PDIP and then recombines the results. The
second uses proximal methods to avoid matrix inversions.

\subsection{Consensus Optimization}
\label{sec:consOpt}

\begin{comment}
General (unconstrained) optimization problems $\min_z
\sum_i f_i(z)$, $z\in\mathbb{R}^n$ can be rewritten equivalently using a
collection of local variables, $x_i\in \mathbb{R}^{n_i}$, and a
constraint. Following the notation of 
\cite{boyd_distributed_2011}, let $k=\mathscr{G}(i,j)$ which
means that the $j^\text{th}$ entry of $x_i$ is $z_k$ (or
$(x_i)_j=z_k$) and define $\tilde{z}_i \in \mathbb{R}^{n_i}$ by
$(\tilde{z}_i)_j=(x_i)_j$. Then the original unconstrained
optimization problem is equivalent to the following constrained
problem
$\min_{\{x_1,...,x_N  \}} \sum_i f_i(x_i)$ subject to $\tilde{z}_i=x_i$.
Now, we can apply ADMM to the augmented Lagrangian of this
problem: 
\begin{align}
x_i&\leftarrow \argmin_{x_i} f_i(x_i) + (u_i)^\top x_i +
     (\rho/2) \norm{ x_i-\tilde{z}_i }_2^2\\
%\label{eq:ADMM_steps}
z_k&\leftarrow (1/S_k)\sum_{\mathscr{G}(i,j)=k} (x_i)_j\\
u_i&\leftarrow u_i + \rho (x_i-\tilde{z}_i)
\end{align}
\end{comment}

Consider an optimization problem of the form $\min_h f(h)$, where
$h\in\mathbb{R}^n$ is the \textit{global variable} and
$f(h):\mathbb{R}^n \rightarrow \mathbb{R}\cup \{+\infty\}$ is
convex. Consensus optimization breaks this problem
into several smaller sub-problems that can be solved independently in
each iteration.

Assume it is possible to define a set of local variables
$x_i \in \mathbb{R}^{n_i}$ such that $f(h)=\sum_i f_i(x_i)$, where
each $x_i$ is a subset of the global variable $h$. More specifically,
each entry of the local variables corresponds to an entry of the
global variable. Therefore we can define a mapping $\mathscr{G}(i,j)$
from the local variable indices into the global variable indices:
$k=\mathscr{G}(i,j)$ means that the $j^\text{th}$ entry of $x_i$ is
$h_k$ (or $(x_i)_j=h_k$). For ease of notation, define $\tilde{h}_i
\in \mathbb{R}^{n_i}$ as $(\tilde{h}_i)_j=h_{\mathscr{G}(i,j)}$. Then,
the original optimization problem is equivalent to the following
problem:  
\begin{equation}
\begin{aligned}
&\min_{\{x_1,...,x_N  \}} \sum_{i=1}^N f_i(x_i) &
s.t. &\quad \tilde{h}_i=x_i.
\end{aligned}
\label{eq:consADMM}
\end{equation}
It is important to note that each entry of the global variable may
correspond to several entries of the local variables and so the
constraints $\tilde{h}_i=x_i$ enforce the consensus between the local
variables corresponding to the same global variable.  
The augmented Lagrangian corresponding to 
\eqref{eq:consADMM} is $L_\rho(x,h,y)=\sum_i
\big(f_i(x_i)+u_i^\top(x_i-\tilde{h}_i) + (\rho/2) \lVert
x_i-\tilde{h}_i \lVert_2^2 \big)$. Now, we can apply ADMM to
$L_\rho$. This results in solving $N$ independent optimization
problems followed by a step to achieve consensus among the solutions
in each iteration. 
% which results in the following ADMM updates:
% \begin{align}
% x_i&\leftarrow\argmin_{x_i} f_i(x_i) + (u_i)^\top x_i +
%      (\rho/2)  \norm{ x_i-\tilde{h}_i }_2^2\\
% h_k&\leftarrow (1/S_k)\sum_{\mathscr{G}(i,j)=k} (x_i)_j \label{eq:ADMM_steps}\\
% u_i&\leftarrow u_i + \rho (x_i-\tilde{h}_i).
% \end{align}
% \attn{If you hate the left arrow, that's fine. But change it
%   everywhere, and use $=$. The $:=$ notation means that ``thing on
%   left is defined to be thing on right'' and isn't really appropriate here.}
% Here, $S_k$ is the number of local variable entries that correspond to
% $h_k$, and $u_i$ are the Lagrange multipliers.  
To solve the optimization problem \eqref{eq:l1tf_var_st_mat} using
this method, we need to address two 
questions: first, how to choose the 
local variables $x_i$, and second, how to the update them.
\begin{figure}[tb]
  \centering
  \includegraphics[width=.6\columnwidth]{Figures/data_cube}
  \caption{The cube represents the global variable $h$ in space and
    time. The sub-cubes specified by the white lines are
    $x_i$.}
  \label{fig:data_cube}
\end{figure}

In \autoref{fig:data_cube}, the global variable $h$ is represented
as a cube. We decompose $h$
into sub-cubes as shown by white lines. Each global variable inside
the sub-cubes corresponds to only one local variable. The global
variables on the border (white lines), however, correspond to more
than one local variable. With 
this definition of $x_i$, the objective
\eqref{eq:l1tf_var_st_mat} decomposes as $\sum_i f_i(x_i)$ where
$f_i(x_i)=-l(y_i\given x_i)+\Lambda_{(i)}^\top |D_{(i)}x_i|$, and
$\Lambda_{(i)}$ and $D_{(i)}$ contain the temporal and spatial
penalties corresponding to $x_i$ only in one sub-cube along with its
boundary. Thus, we now need to use PDIP to solve $N$ problems each of
size $n_i$, which is feasible for small enough
$n_i$. \autoref{alg:conADMM} gives the general version of this
algorithm. A more detailed discussion of this is in Appendix B of the
Supplement where we show how to compute the dual and the derivatives
of the augmented Lagrangian.

\begin{algorithm}[tb]
  \caption{Consensus ADMM }
  \label{alg:conADMM}
  \begin{algorithmic}[1]
    \STATE {\bfseries Input:} data $y$, penalty matrix $D$, 
    $\epsilon, \rho,\lambda_t,\lambda_s >0$.
    \STATE {\bf Set:} $h\leftarrow 0$, $z\leftarrow 0$, $u\leftarrow
    0$. \COMMENT{Initialization} 
    \REPEAT
    \STATE $\begin{aligned}x_i&\leftarrow\argmin_{x_i} -l(y_i\given
    x_i)+\Lambda_{(i)}^\top |D_{(i)}x_i|\\&\quad+ (u_i)^\top x_i +
    (\rho/2)  \lVert x_i-\tilde{h}_i \rVert_2^2\end{aligned}$. \\\COMMENT{Update local
      vars using PDIP}
    \STATE $h_k\leftarrow (1/S_k)\sum_{\mathscr{G}(i,j)=k} (x_i)_j
    $. \COMMENT{Global update.}
    \STATE $ u_i\leftarrow u_i + \rho (x_i-\tilde{h}_i)$. \COMMENT{Dual update}
    \UNTIL {$\max\left\{\norm{h^{m+1}-h^m},\ \norm{h^m-x^m}\right\} < \epsilon$}
    \STATE {\bf Return:} $h$.
  \end{algorithmic}
\end{algorithm}

Because consensus ADMM breaks the large optimization into
sub-problems that can be solved independently, it is amenable to a
split-gather parallelization strategy via, e.g., the map reduce framework.
In each iteration, the
computation time will be equal to the time to solve each sub-problem
plus the time to communicate the solutions on the master processor
and perform the consensus step. Since each sub-problem is
small, with parallelization, the computation time in each iteration
will be small. In addition, our experiments with several values of
$\lambda_t$ and $\lambda_s$ showed that the algorithm converges in a few
hundred iterations. 
% \attn{Need to redo this:}Solving each sub-problem on a machine with four
% 3.20GHz Intel i5-3470 cores takes less than 3 seconds on average, and
% so for example if we assume that communication time is 10 seconds and
% the algorithm converges in 300 iterations, with parallelization on
% $N_{sub-cubes}$ machines, the algorithm will converge in about 1
% hour. Assuming that we use $N_{sub-cubes}$ machines and that the
% convergence rate of the algorithm is independent of the grid size,
% this time will be independent of the grid size. 
% If we perform these computations on a single machine, the computation
% time grows linearly with $N_{sub-cubes}$. For example, for the data in
% a grid over the united states and using $3\times3\times521$ sub-cubes
% each iteration of the algorithm will take about 20 minutes on a single
% machine and so with 300 iterations it will take several days to
% converge. Given that we need to compute the solution for several
% values of the parameters $\lambda_t$ and $\lambda_s$, this computation
% time is not feasible. 
This algorithm is most useful if we can parallelize the
computation over several machines with low communication cost between
machines. In the next section, we describe 
another algorithm which makes the computation feasible on a single
machine.

\subsection{Linearized ADMM}
\label{sec:linADMM}

% In this section, we describe \textit{Linearized ADMM algorithm} \cite{parikh_proximal_2014} which, as we will see, makes the computation on a single machine feasible.

Consider the generic optimization problem
$\min_x f(x)+g(Dx)$
where $x\in \mathbb{R}^n$ and $D\in \mathbb{R}^{m\times n}$. Each
iteration of the linearized ADMM
algorithm~\cite{parikh_proximal_2014} for solving this problem 
has the form
\begin{equation}
  \label{eq:linADMM_steps}
  \begin{aligned}
    x & \leftarrow \prox_{\mu f} \left(x - (\mu/\rho)D^\top (D x - z + u )\right)\\
    z & \leftarrow \prox_{\rho g} \left(z + u\right)\\
    u & \leftarrow u + D x - z
  \end{aligned}
\end{equation}
where the algorithm parameters $\mu$ and $\rho$ satisfy $0 < \mu <
\rho/\norm{D}_2^2$, $z,u\in \mathbb{R}^m$ and the proximal operator is
defined as $\prox_{\alpha \varphi}(u) = \min_x \,\, \alpha \cdot
\varphi(x)+\frac{1}{2} \norm{ x-u}_2^2$. Proximal algorithms are feasible
when these proximal operators can be evaluated efficiently which, as
we show next, is the case.

%Clearly, \eqref{eq:l1tf_var_st} has this form necessary for using this algorithm.
% The optimization problem 
% form \eqref{eq:linADMM_opt} by defining
% \begin{align}
% & f(x):= \sum_{k} f_k(x_k) := \sum_{k}x_k+y_{k}^2e^{-x_{k}}\\
% & g(z):= \sum_{l} g_l(z_l) := \sum_{l} \lambda_l |z_l|\label{eq:linADMM_fg}\\
% & z = Dx
% \end{align}
% where $y_k$ is the $k^{th}$ entry of the vector whose entries are
% $y_{ijt}$, and $\lambda_l$ is the $l^{th}$ entry of the vector
% $\Lambda^\top=(\lambda_t\textbf{e}_{n_t}^\top|\lambda_s\textbf{e}_{n_s}^\top)$
% (see \autoref{sec:exten}). 
% To perform the steps in \eqref{eq:linADMM_steps}, we need to evaluate
% $\prox_{\mu f}$ and $\prox_{\rho g}$.

>>>>

\documentclass[letterpaper]{article} %DO NOT CHANGE THIS
\usepackage{aaai19}  %Required
\usepackage{times}  %Required
\usepackage{helvet}  %Required
\usepackage{courier}  %Required
\usepackage{url}  %Required
\usepackage{graphicx}  %Required
\frenchspacing  %Required
\setlength{\pdfpagewidth}{8.5in}  %Required
\setlength{\pdfpageheight}{11in}  %Required
%PDF Info Is Required:

\pdfinfo{
  /Title (Supplement for Algorithms for Estimating Trends in Global Temperature
  Volatility )
   %/Author (Khodadadi and McDonald)
  /Author (John Doe, Jane Doe)}

@@

\usepackage{gensymb}
\usepackage{xcolor}
\newcommand{\attn}[1]{\textcolor{red}{TODO: #1}}
\newcommand{\one}{\mathbf{1}}
\renewcommand{\algorithmiccomment}[1]{\hfill $\rhd$ #1}
\newcommand{\given}{\;\vert\;}
\newcommand*{\algorithmautorefname}{Algorithm}%
\newcommand{\norm}[1]{\left\lVert #1 \right\rVert}
\newcommand{\y}{\textrm{year}}
\DeclareMathOperator*{\prox}{\bf prox}
\DeclareMathOperator*{\sign}{sign}

\usepackage{xr}
\externaldocument{khodadadi-mcdonald-aaai19}

@@

\title{Supplement for Algorithms for Estimating Trends in Global Temperature Volatility}

@@
